Unsupervised Characterization of Water Composition with UAV-based Hyperspectral Imaging and Generative Topographic Mapping

### 2024-06-19 paper/paper.tex
Updates from Overleaf

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Unmanned Aerial Vehicles (UAVs) equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies. The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods. However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain. In this study, we demonstrate how the Generative Topographic Mapping (GTM), a probabilistic realization of the Self-organizing Map, can be used to visualize high-dimensional hyperspectral imagery and extract spectral signatures corresponding to unique endmembers present in the water.  Using data collected across a North Texas pond, we first apply the GTM to visualize the distribution of captured reflectance spectra revealing small-scale spatial variability of water composition. Next, we demonstrate how the nodes of the fitted GTM can be interpreted as unique spectral endmembers. Using extracted endmembers together with the normalized spectral similarity score, we are able to efficiently map the abundance of near shore algae as well as the evolution of a rhodamine tracer dye used to simulate water contamination by a localized source. 
%This study highlights the value of combining UAV-based hyperspectral imaging with unsupervised methods for...
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality. Using data collected at a pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI. We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection. Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including land pixels, algae, water, and a simulated contaminant plume using a rhodamine tracer dye. Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond. We demonstrate this capability by using a GTM to map the abundance of algae near the shore as well as the dispersion of a rhodamine dye plume.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

\begin{figure}[t]
%\begin{adjustwidth}{-\extralength}{0cm}
\centering
%\includegraphics[width=15.5cm]{paper/figures/methods/sample-spectra.pdf}
\includegraphics[width=\columnwidth]{paper/figures/methods/sample-spectra.pdf}
%\end{adjustwidth}
\caption{(\textbf{a}) Sample hyperspectral data cube. Spectra are plotted using their geographic position with the log10-reflectance colored along the z axis and a pseudo-color image on top. (\textbf{b}) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and dry grass. (\textbf{c}) Reflectance spectra for the exemplar points scaled so the peak value of each spectrum is $1.0$. (\textbf{d}) The location of the pond in Montague, Texas where data were collected for this study.\label{fig:sample-spectra}}
\end{figure}

In this study, we explore the use of the GTM as a tool for dimensionality reduction and non-linear endmember extraction of hyperspectral imagery. To this end, a dataset of HSI were collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}. In this section we first provide a detailed overview of the GTM algorithm. Next we describe the UAV platform used for HSI collection and the various steps used to processes captured HSI. Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.

@@

After conversion to reflectance, each HSI must also be georectified to assign geographic coordinates to each pixel. The UAV flights were performed at an approximate altitude of $50$ m above the water so that the imaging surface can be considered to be flat. Consequently, HSI were rapidly georectified to a $10$ cm resolution using position and orientation data from the embedded GPS/INS as outlined in \cite{muller2002program, baumker2001new, mostafa2000multi}. An example hyperspectral data cube is illustrated in panel (a) of Fig~\ref{fig:spectral-signatures} where the log10-reflectance values are plotted along the z-axis for each pixel in the scene and a psuedo-color image at the top of the data cube illustrates how the water would appear to the human eye from the perspective of the UAV.

@@

To explore the ability of the GTM to segment HSI pixels and aid in source identification we first consider a dataset of water-only spectra consisting of $>36,000$ records identified from collected HSI by a normalized difference water index (NDWI) greater than $0.25$ where the NDWI is defined as 
\begin{equation}
    \text{NDWI} = \dfrac{R(550) - R(860)}{R(550) + R(860)}.
\end{equation}
We then apply the trained GTM to all water-only pixels to visualize the distribution learned by the GTM and examine spectra associated with a subset of nodes in the latent space in order to assess the small-scale spatial variability within the pond.

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ pixels were sampled for model training. Exemplar spectra for water, grass, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}. Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories. Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

@@

A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Sec~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI. The resulting GTM is visualized in Fig~\ref{fig:gtm-water}. First, we note that the GTM has utilized most of the latent space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted. For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the latent space. Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
%\caption{GTM Map: We plot the mean position of each data point in the latent space with points colored by the their associated NDVI values. The locations of the exemplar points for algae, rhodamine dye, water, and dry grass are shown in the latent space illustrating that the GTM has clearly distinguished land, near-land, and water pixels.\label{fig:gtm-ndvi}}
\caption{GTM latent space visualization: Each sample spectrum from the dataset of combined HSI covering shore, water, and the rhodamine dye release are plotted in the GTM latent space at the location of the mean node responsibility, $\hat{\xi}_n$. Points are colored according to the NDVI computed from the original reflectance spectra. The locations of exemplar spectra for algae, rhodamine, water, and grass in the latent space are included as color-filled circles. Spectra from the shore and water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
\end{figure}

A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release. The resulting class map is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and grass are included as color-filled circles further illustrating the classification obtained by the GTM.

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/spectral-signatures.png}
\end{adjustwidth}
\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), dry grass (bottom left), and open water (bottom right). A pseudo-color image is inset into each plot with the location of the exemplar spectrum marked with a white circle. \label{fig:spectral-signatures}}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

@@

The second application of the GTM presented in this study is for the extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, grass, and water demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Unmanned Aerial Vehicles equipped with hyperspectral imagers have emerged as an essential technology for the characterization of inland water bodies. The high spectral and spatial resolutions of these systems enable the retrieval of a plethora of optically-active water quality parameters via band ratio algorithms and machine learning methods. However, fitting and validating these models requires access to sufficient quantities of in situ reference data which are time-consuming and expensive to obtain. In this study, we demonstrate how the Generative Topographic Mapping (GTM), a probabilistic realization of the Self-organizing Map, can be used to visualize high-dimensional hyperspectral imagery and extract spectral signatures corresponding to unique endmembers present in the water.  Using data collected across a North Texas pond, we first apply the GTM to visualize the distribution of captured reflectance spectra revealing small-scale spatial variability of water composition. Next, we demonstrate how the nodes of the fitted GTM can be interpreted as unique spectral endmembers. Using extracted endmembers together with the normalized spectral similarity score, we are able to efficiently map the abundance of near shore algae as well as the evolution of a rhodamine tracer dye used to simulate water contamination by a localized source. 
%This study highlights the value of combining UAV-based hyperspectral imaging with unsupervised methods for...
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{introduction}

@@

In this paper, we explore the application of the GTM to UAV-acquired HSI for the characterization water quality. Using data collected at a pond in Montague, North Texas, we first train a GTM using water-only pixels to produce a low-dimensional representation of the collected HSI. We use this mapping to explore the highly detailed small-scale variability within the pond and discuss how this can be used to guide reference data collection. Next, we demonstrate how the GTM can be utilized to identify relevant spectral endmembers from a combined dataset including land pixels, algae, water, and a simulated contaminant plume using a rhodamine tracer dye. Once identified, these endmembers can be used to rapidly map the abundance of spectral features within the pond. We demonstrate this capability by using a GTM to map the abundance of filamentous blue-green algae near the shore as well as the dispersion of a rhodamine dye plume.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}\label{materials-and-methods}

\begin{figure}[t]
%\begin{adjustwidth}{-\extralength}{0cm}
\centering
%\includegraphics[width=15.5cm]{paper/figures/methods/sample-spectra.pdf}
\includegraphics[width=\columnwidth]{paper/figures/methods/sample-spectra.pdf}
%\end{adjustwidth}
\caption{(\textbf{a}) Sample hyperspectral data cube. Spectra are plotted using their geographic position with the log10-reflectance colored along the z axis and a pseudo-color image on top. (\textbf{b}) Points taken from a sample hyperspectral data cube corresponding to algae, rhodamine dye, water, and ground (dirt and dry grass). (\textbf{c}) Reflectance spectra for the exemplar points scaled so the peak value of each spectrum is $1.0$. (\textbf{d}) The location of the pond in Montague, Texas where data were collected for this study.\label{fig:sample-spectra}}
\end{figure}

In this study, we explore the use of the GTM as a tool for dimensionality reduction and unsupervised endmember extraction of hyperspectral imagery. To this end, a dataset of HSI were collected at a pond in Montague, North Texas (shown in panel d of Fig~\ref{fig:sample-spectra}) on 23 November 2020 using a UAV mounted hyperspectral imager configured as described in \cite{robot-team-1, robot-team-2}. In this section we first provide a detailed overview of the GTM algorithm. Next we describe the UAV platform used for HSI collection and the various steps used to processes captured HSI. \textcolor{red}{Finally, we describe the two case studies presented in this paper for the utilization of the GTM as a HSI visualization tool and as an endmember extraction technique.}

@@

After conversion to reflectance, each HSI must also be georectified to assign geographic coordinates to each pixel. The UAV flights were performed at an approximate altitude of $50$ m above the water so that the imaging surface can be considered to be flat. Consequently, HSI were rapidly georectified to a $10$ cm resolution using position and orientation data from the embedded GPS/INS as outlined in \cite{muller2002program, baumker2001new, mostafa2000multi}. An example hyperspectral data cube is illustrated in panel (a) of Fig~\ref{fig:sample-spectra} where the log10-reflectance values are plotted along the z-axis for each pixel in the scene and a psuedo-color image at the top of the data cube illustrates how the water would appear to the human eye from the perspective of the UAV.

@@

To explore the ability of the GTM to segment HSI pixels and aid in source identification we first consider a dataset of water-only spectra consisting of $>36,000$ records identified from collected HSI by a normalized difference water index (NDWI) greater than $0.25$ where the NDWI is defined as 
\begin{equation}
    \text{NDWI} = \dfrac{R(550) - R(860)}{R(550) + R(860)}
\end{equation}
where $\lambda=550$ and $\lambda=860$ nm were chosen to represent the green and NIR bands as described in \cite{ndwi}. We then apply the trained GTM to all water-only pixels to visualize the distribution learned by the GTM and examine spectra associated with a subset of nodes in the latent space in order to assess the small-scale spatial variability within the pond.

Next, we consider a dataset of combined HSI pixels including both water and land. To simulate the dispersion of a potential contaminant source, a Rhodamine tracer dye was released into the western portion of the pond and two additional UAV flights were used to collect HSI capturing the evolution of the resulting plume. From these HSI a collection of $>145,000$ pixels were sampled for model training. Exemplar spectra from water, ground, algae, and rhodamine dye were identified from a sample HSI as shown in panels (b) and (c) of Fig~\ref{fig:sample-spectra}. Using these spectra, we explore the trained GTM and use it to extract spectral signatures corresponding to these endmember categories. Values for the hyperparameters $m$, $s$, and $\alpha$, are determined by training multiple GTM models and selecting values which minimize the Bayesian Information Criterion (BIC) given by 
\begin{equation}
    \text{BIC} = P\ln(N) - 2\mathcal{L}
\end{equation}
where $P$ is the total number of model parameters, $N$ is the number of records in the dataset, and $\mathcal{L}$ is the log-likelihood defined in Eq~\ref{eq:llh}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}\label{results}

@@

A GTM with $K=32\times 32$ nodes was trained on the dataset of water-only pixels described in Section ~\ref{sec:case-studies} in order to explore the distribution of reflectance spectra captured by the HSI. The resulting GTM is visualized in Fig~\ref{fig:gtm-water}. First, we note that the GTM has utilized most of the latent space as illustrated panel (b) where the position corresponding to the mean responsibility of each data point $\hat{\xi}_n$ has been plotted. For this water-only GTM, the spectra appear to be largely clustered toward the left edge, bottom, and right edge of the latent space. Spectral signatures corresponding to GTM nodes from the four corners and center of the latent space are shown in panel (c) illustrating the spectral variability represented across the latent space.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.8\columnwidth]{paper/figures/results/square-ndvi.pdf}
\caption{GTM latent space visualization: Each sample spectrum from the dataset of combined HSI covering shore, water, and the rhodamine dye release are plotted in the GTM latent space at the location of the mean node responsibility, $\hat{\xi}_n$. Points are colored according to the NDVI computed from the original reflectance spectra. The locations of exemplar spectra for algae, rhodamine, water, and ground points in the latent space are included as color-filled circles. Spectra from the shore and water are clearly separated to different regions of the latent space.\label{fig:gtm-ndvi}}
\end{figure}

A second GTM was trained on the combined dataset of HSI covering shore, water, and the rhodamine tracer dye release. The resulting class map is shown in Fig~\ref{fig:gtm-ndvi} where each sample has been plotted at the position of the mean node responsibility and colored by the normalized difference vegetation index (NDVI) computed from the original reflectance spectrum. The NDVI is a spectral index sensitive to variations in vegetation health with negative values corresponding to water, small positive values corresponding to sparse vegetation, and large values near $1$ corresponding to dense vegetation \cite{thenkabail2018hyperspectral}. From this, we see that the GTM can clearly separate spectra by vegetation content with high values concentrated to the left of the latent space and negative, water-based pixels concentrated to the right. Additionally, the position of exemplar spectra for algae, rhodamine, water, and ground points are included as color-filled circles further illustrating the classification obtained by the GTM.

@@

\begin{figure}[t]
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=17.0cm]{paper/figures/results/spectral-signatures.png}
\end{adjustwidth}
\caption{Spectral signatures $\psi(\xi_k)$ corresponding to GTM nodes with non-zero responsibility for exemplar spectra corresponding to the rhodamine dye plume (top left), algae (top right), ground (bottom left), and open water (bottom right). A pseudo-color image is inset into each plot with the location of the exemplar spectrum marked with a white circle. \label{fig:spectral-signatures}}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}\label{discussion}

@@

The second application of the GTM presented in this study is for the unsupervised extraction of endmembers corresponding to unique sources observed in the HSI. Here the GTM is an attractive choice as it does not rely on the assumption of linear mixing and the presence of pure pixels which are easily broken by the effects of multiple scattering in realistic scenarios \cite{nonlinearity-in-hsi}. Additionally, because the GTM is a probabilistic model, the values of model hyperparameters can be selected objectively using information criteria like the BIC. This is a clear advantage over similar methods like the SOM on which the GTM is based. Endmembers identified for exemplar spectra corresponding to rhodamine, algae, ground, and water pixels demonstrate that the method can successfully identify spectral signatures corresponding to diverse sources. If a set of labeled spectra for known sources are available, their representation in the GTM latent space could be used to perform as semi-supervised classification similar to the SOM approach developed by Riese et al. \cite{riese2019supervised}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}\label{conclusions}
